Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-17 conclusion/conclusion.tex
New fix
Grammar fix

 To conclude, ABC SMC was successfully applied on the models of zebrafish spinal cord regeneration. These models were all ODE equations that described the interactions and effects between cells and proteins in the lesion site. These equations contains four dependent variable and more than 10 model parameters.  Parameter estimation for high-dimensional parameter space is a challenging work for many dynamic system models, as insufficient explore of parameter space can easily lead to local optimal. Using a large population size in ABC SMC in our models could partially relief the concerns in local modes, and give considerable parameter estimates (FIGURE). Also ABC SMC successfully helped in the model comparison and thus can be used for suggestions of possibly best model, although there were uncertainties remained in the process and a large number of population size was need.

 ABC SMC implementation is a complex process with respect to both algorithm settings and code development. The ability of parameter inference and robustness of algorithm were firstly studied using synthetic data where the true parameter value was known, and experiments were performed to find proper hyperparameter and implementation settings of ABC SMC. Later inference on the observed data followed the suggestions from the synthetic data experiments and acceptable results were produced.

 The inference program was mostly run on multi-core machines, and additional scaling-up performance test illustrated the scaling behaviour. When the number of running \verb|Python| process exceeds the number of physical cores, hyperthreading would be enabled but gains not significant improvements with respect to execution time, while increase of average sampling size per second was still observed.

@@

 From this application, some suggestions on ABC SMC implementation could be drawn. How to set the prior could be the most tricky decision for a high dimensional parameter inference. The distribution and the interval of each parameter can affect the inference results significantly and improper prior cold lead to poor fit. The epsilon trajectory and joint distribution of parameters can be used to identity possible local modes; if stuck, it takes much more samples than usual to jump out out local optimal. To avoid this a large population size and a proper epsilon schedule may help.for the uncertainties observed in model selection and some duplicated runs with high variance, a longer run with more generation could help to identify whether the convergency is consistent such that a reproducible inferred results can be obtained.

@@

In the aspect of models, some further models can be studied. If there were some more experimental evidence on the proposed interaction map, we could proposed more precise models, or calibrate some terms in our existing models. Also, models of other forms, e.g. PDE and stochastic models, could also be studied if a more complex model is needed for the dynamic system. Simplification is also possible of existing models

>>>>

 To conclude, ABC SMC was successfully applied to the models of zebrafish spinal cord regeneration. These models were all ODE equations that described the interactions and effects between cells and proteins in the lesion site. These equations contain four dependent variable and more than 10 model parameters.  Parameter estimation for high-dimensional parameter space is a challenging work for many dynamic system models, as insufficient explore of parameter space can easily lead to local optimal. Using a large population size in ABC SMC in our models could partially relief the concerns in local modes, and give considerable parameter estimates. We found that the first three models could not recover some local features observed in the early times, so further models -- model 4 and 5 -- were proposed. Additional interactions were introduced in the new model, and the results proved that they were helpful in resolving the under-fit.

 Also, ABC SMC successfully helped in the model comparison and thus can be used for suggestions of the best model, although there were uncertainties remained in the process and a large number of population size needed. In our practice, model 5 won with absolute advantages when the threshold is converged to a small value. The approximated posteriors showed bell-shaped density distribution for most of the parameters (Figure \ref{fig:model5_para} and Figure \ref{fig:para1}), indicating that most parameters were well-inferred in our implementations.

%  ABC SMC implementation is a complex process with respect to both algorithm settings and code development. The ability of parameter inference and robustness of algorithm was firstly studied using synthetic data where the true parameter value was known, and experiments were performed to find proper hyperparameter and implementation settings of ABC SMC.

% Later inference on the observed data followed these suggestions from the synthetic data experiments and acceptable results were produced. These examples proved that ABC SMC was capable of the inference tasks for the proposed models with high-dimensional parameter space.

 The inference program was mostly run on multi-core machines, and additional scaling-up performance test illustrated the scaling behaviour. When the number of running \verb|Python| process exceeds the number of physical cores, hyperthreading would be enabled, and performance improvement was still considerable (Figure \ref{fig:sample_per_sec}). Studies on the variant required samples revealed that some threshold paths could lead to local optimum and significantly affect the efficiency. A non-linear relationship between the required samples and execution time was also observed in the first two generations.

@@

 From this application, some suggestions on ABC SMC implementation could be drawn. How to set the prior could be the most tricky decision for high dimensional parameter inference. The distribution and the interval of each parameter can affect the inference results significantly, and improper prior cold lead to poor fit. In our cases, log-uniform prior gave better fits than uniform prior. The threshold path and plot of required samples in each generation can be used to identify possible local modes; if stuck, it takes much more samples than usual to move out. To avoid this, large population size and a proper epsilon schedule may help. For the uncertainties observed in model selection and some duplicated runs with high variance, a longer run with more generation could help to identify whether the convergency is consistent such that a reproducible inferred results can be obtained.

@@

In the aspect of models, some further models can be studied. If there were some more experimental evidence on the proposed interaction map, we could propose more precise models, or calibrate some terms in our existing models. Also, models of other forms, e.g. PDE and stochastic models, could also be studied if a more complex model is needed for the dynamic system. Simplification is also possible of existing models. Some highly-correlated parameters could be reduced, as there exist some linear relationships between parameters.

% Beyond this scope, we are interested in the

### 2020-08-17 implementation/implementation.tex
Grammar fix

Model 4 did not overweight other models across all the generations; model 3 gained an advantage before generation 15 except generation 5. After generation 18 ($\epsilon_{18} = 16.98$), model 5 gained absolute advantage and won in further generations. Also, repeated runs showed similar trends: although uncertainties were present before generation 20, model 5 won with nearly 100\% probability in later generations where epsilon values were considerably low, i.e. $\epsilon_t<16$.

>>>>

Model 4 did not overweight other models across all the generations; model 3 gained an advantage before generation 15 except generation 5. After generation 18 ($\epsilon_{18} = 16.98$), model 5 gained absolute advantage and won in further generations. Also, repeated runs showed similar trends: although uncertainties (e.g. violently changes mode probabilities) were present before generation 20, model 5 won with nearly 100\% probability in later generations where epsilon values were considerably low, i.e. $\epsilon_t<16$.

### 2020-08-17 performance/performance.tex
Grammar fix

The performance experiments were designed to explore the parallel performance of ABC SMC inference framework which were used in the parameter estimation and model selection tasks. Usually ABC SMC is a time-consuming and computation intensive task and ideally executed on large clusters. The scheduling strategy, implementation details and many other factors can affect the parallel efficiency.

@@

First experiments are designed to illustrate the scaling-up performance. The program used here is an implementation of ABc SMC on model 5. The details of the ABC SMC settings is listed below

@@

For HPC systems like Cirrus, \verb|pyabc| uses \verb|multiprocessing| for multi-core parallel sampling. By default if the number of cores is not specified, it will automatically read the number of available cores and use them all. Cirrus has a 36-core CPU which support hyperthreading, such that the maximal number of cores available to \verb|multiprocessing| is 72.

The program is executed on Cirrus, using 8, 16, 24, 36, 54 and 72 cores respectively. Each run is repeated 10 times. The average execution time, required sampling numbers are recorded. Hyperthreading is enabled when using 54 and 72 cores. The access to the node that contains computation cores is exclusive, such that the execution would not be affected by other programs of operations.

The implementation of ABC SMC in \verb|pyabc| enables the parallelisation of sampling, which is the must time-consuming part. The rest part of the program is mostly not parallelised, e.g. database I/O and reductions operations. The sampling process involves sampling, perturbation and test of the acceptance criteria, all of which are computation-intensive.

In practice, using ABC SMC to estimate the parameters of a given model could cost up to several hundred of hours if the computational resources is limited[REF]. The performance experiment result could provide a reference that illustrate that how the efficiency changes when scaling-up or the trade-offs in computational resources' cost and their benefit.

@@

The recorded execution time using different number of cores were used to calculate the speedup of the program (Figure \ref{fig:speedup}). Here 8 cores is regarded as the baseline for speed-up and efficiency, as the serial version (using only one core) can take quite a long time to finish 10 repeats. When increasing number of cores, the execution time drops fast at first but the decreasing rate (dropping speed, i.e. the derivative of the execution time against time) is gradually reduced to zero; 36, 54 and 72 cores gives nearly the same average execution time and consequently gains very close speedup values (Table \ref{table:performance}).

@@

Denote the number of cores used in the ABC SMC runs as variable $p$. The speedup curve shows a linear trend when $p\leq 36$; when $p\geq 36$ the speedup stays nearly unchanged. $p=36$ is an inflection point, where 36 is the maximum numbers available physical cores in a compute node of Cirrus. A constant drop of efficiency is observed when increasing $p$; smaller $p$ ($p<36$) gives an efficiency higher than 65\% but greater $p$ e.g. $p=72$ results in a low efficiency (34\%).

@@

Due to the variant required samples, our interests switched to the per-second performance, where the average sampling numbers per second under different of number of cores ($p$) are plotted (Figure \ref{fig:sample_per_sec}). The plot shows that even theoretical speedup stops growing when $p>36$, the average sampling numbers per second performance still get improved when using more than 36 cores. It proves that the implementation used here scales reasonably well without encountering bottleneck when using multi-core parallelisation inside one physical node.

The variant total required samples for different runs under the same setting was most likely the result of different threshold evolve path. Median epsilon schedule and randomness in the sampling process resulted in different sequences of threshold for different runs, which means different repeated runs may have different convergency path \cite{threshold}. For example, Some runs temporarily converges to a local optimum, while the target epsilon in the next generation cannot possibly be reached if the new particles are still concentrated in the local optimum. In this case, huge amount of sample will be drawn to find enough accepted particles, and new generation will be able to jump out of the local optimum and have different concentrations.

@@

To illustrate this, Two independent run, namely Run A and Run B, were chosen from the experiment of $p=24$ and visualised in Figure \ref{fig:local_modes} and Figure \ref{fig:local_para}. Thess two run used identical options of the algorithm. Figure \ref{fig:local_modes} (a) shows the comparison of the required samples, where Run A converged to a local mode, and in population 11 took much more sample tires to move away from the local optimum (Figure \ref{fig:local_modes} (b)). Figure \ref{fig:local_para} compares the posterior density distribution before (t=9, cyan) and after (t=13, red) the jumping away: significant movement of concentrations in posterior density distributions are observed, especially in parameters from ODE 1, 2 and 4. An interning movement is that there are two concentrations in $\kappa_{\Phi\beta}$ and $\mu_\Phi$ when t=9, and the majority of particles are concentrated in the second concentration with high parameter values; when t=13, the second concentration distribution disappears and all particles are concentrated near zero.

@@

Moreover, we found that the execution time was not ideally proportional to the sampling tires in that period of time, for both different repeated runs and different generations in a single run. From the same example, Figure \ref{fig:local_modes} (b) and (c) show the non-linear relationship between execution time and sample tries in different generations. Apart from 1st and 2nd generation, generation 3 to 20 show a general linear relationship i.e. $T_t\propto n_t$, where $t$ is the population index, $T_t$ is the execution time of generation $t$ and $n_t$ is the required samples in generation $t$. However, this relationship does not hold for the first two generation. Generation 1 is to draw sample from prior, but it is done with high efficiency as only small amount of time are taken; generation 2 takes unusual large amount of the, but it does not require much more samples. The abnormal execution time of generation 1 and two are observed in all other runs, where for $p=24$ generation 1 usually takes c. 16s and generation 2 takes c. 1600s. This is believed to be related to the initialisation of the algorithm.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited the the sampling, calculation and criterion test process; some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. Database I/O is also serial.

>>>>

The performance experiments were designed to explore the parallel performance of ABC SMC inference framework that was used in our parameter estimation and model selection tasks. Generally, ABC SMC is a time-consuming and computation-intensive task and ideally executed on large clusters. The scheduling strategy, implementation details, randomness in the algorithm and many other factors can affect the parallel efficiency. Thus, we designed a scaling-up experiment and studied a case where abnormal performance was observed.

@@

First experiments are designed to illustrate the scaling-up performance. The program used here is an implementation of ABC SMC on model 5. The details of the ABC SMC settings are listed below

@@

For HPC systems like Cirrus, \verb|pyabc| uses \verb|multiprocessing| for multi-core parallel sampling. By default, if the number of cores is not specified, it will automatically read the number of available cores and use them all. Cirrus has a 36-core CPU which support hyperthreading, such that the maximal number of cores available to \verb|multiprocessing| is 72.

The experiment was executed on Cirrus, using 8, 16, 24, 36, 54 and 72 cores respectively. Each run was repeated ten times and the average execution time, required sampling numbers are recorded. Hyperthreading was enabled when using 54 and 72 cores. The access to the node that contains computation cores is exclusive, such that the execution would not be affected by other programs of operations.

The implementation of ABC SMC in \verb|pyabc| enables the parallelisation of sampling, which is the most time-consuming part. The rest part of the program is mostly not parallelised, e.g. database I/O and reductions operations. The sampling process involves sampling, perturbation and test of the acceptance criteria, which are computation-intensive.

In practice, using ABC SMC to estimate the parameters of a given model could cost up to several hundred of hours if the computational resources are limited \cite{ref:compare}. The performance experiment result could provide a reference that illustrates how the efficiency changes when scaling-up or the trade-offs in computational resources' cost and their benefit.

@@

The recorded execution time under different numbers of cores was used to calculate the speedup of the program (Figure \ref{fig:speedup}). Here 8-core run is regarded as the baseline for speedup and efficiency, as the serial version (using only one core) can take quite a long time to finish 10 repeats. When increasing number of cores, the execution time drops fast at first, but the decreasing trend became slower for large numbers of cores; 36, 54 and 72 cores gives nearly the same average execution time and consequently gains very close speedup values (Table \ref{table:performance}).

@@

Denote the number of cores used in the ABC SMC runs as variable $p$. The speedup curve shows a linear trend when $p\leq 36$; when $p\geq 36$ the speedup stays nearly unchanged. $p=36$ is an inflexion point, where 36 is the maximum numbers available physical cores in a compute node of Cirrus. A constant drop of efficiency is observed when increasing $p$; smaller $p$ ($p<36$) gives an efficiency higher than 65\% but greater $p$ e.g. $p=72$ results in a low efficiency (34\%).

@@

Due to the variant required samples, our interests switched to the per-second performance, where the average sampling numbers per second under different of numbers of cores ($p$) are plotted (Figure \ref{fig:sample_per_sec}). The plot shows that even theoretical speedup stops growing when $p>36$, the average sampling numbers per second performance still get improved when using more than 36 cores. It proves that the implementation used here scales reasonably well without encountering bottleneck when using multi-core parallelisation inside one physical node.

The variant total required samples for different runs under the same setting was most likely the result of different threshold evolve path. Median epsilon schedule and randomness in the sampling process resulted in different sequences of thresholds for different runs, which means different repeated runs may have different convergency path \cite{threshold}. For example, Some runs temporarily converge to a local optimum, while the target epsilon in the next generation cannot possibly be reached if the new particles are still concentrated in the local optimum. In this case, a huge amount of sample will be drawn to find enough accepted particles, and the new generation will be able to jump out of the local optimum and have different concentrations.

@@

To illustrate this, Two independent run, namely Run A and Run B, were chosen from the experiment of $p=24$ and visualised in Figure \ref{fig:local_modes} and Figure \ref{fig:local_para}. These two runs used identical options of the algorithm. Figure \ref{fig:local_modes} (a) shows the comparison of the required samples, where Run A converged to a local mode, and in population 11 took much more sample tires to move away from the local optimum (Figure \ref{fig:local_modes} (b)). Figure \ref{fig:local_para} compares the posterior density distribution before (t=9, cyan) and after (t=13, red) the jumping away: significant movement of concentrations in posterior density distributions are observed, especially in parameters from ODE 1, 2 and 4. An interning movement is that there are two concentrations in $\kappa_{\Phi\beta}$ and $\mu_\Phi$ when t=9, and the majority of particles are concentrated in the second concentration with high parameter values; when t=13, the second concentration distribution disappears and all particles are concentrated near zero.

@@

Moreover, we found that the execution time was not ideally proportional to the sampling tires in that period of time, for both different repeated runs and different generations in a single run. From the same example, Figure \ref{fig:local_modes} (b) and (c) show the non-linear relationship between execution time and sample tries in different generations. Apart from 1st and 2nd generation, generation 3 to 20 show a general linear relationship i.e. $T_t\propto n_t$, where $t$ is the population index, $T_t$ is the execution time of generation $t$ and $n_t$ is the required samples in generation $t$. However, this relationship does not hold for the first two generations. Generation 1 is to draw samples from prior, but it is done with high efficiency as only a small amount of time are taken; generation 2 takes an unusually large amount of the, but it does not require much more samples. The abnormal execution time of generation 1 and two are observed in all other runs, where for $p=24$ generation 1 usually takes c. 16s and generation 2 takes c. 1600s. This is believed to be related to the initialisation of the algorithm.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited to the sampling, calculation and criterion test process. Some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. The database I/O is also serial.
